DySECT -- truly space efficient dynamic hashing

### 2017-04-18 paper/paper.tex
Work - paper (pre Sebastians stuff) plotting (v1.0) fixed ripprob

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Dynamic Space Efficient Cuckoo Table -- DySECT when doubling is too much!} % -- a novel approach to growing!}
\titlerunning{A Sample LIPIcs Article} %optional, in case that the title is too long; the running title should fit into the top page column

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
\author[1]{Tobias Maier}
\author[1]{Dr. Peter Sanders}
\affil[1]{Karlsruhe Institute of Technology, Karlsruhe, Deutschland\\
  \texttt{\{t.maier, sanders\}@kit.edu}}
%% \affil[2]{Department of Informatics, Dummy College, Address/City, Country\\
%%   \texttt{access@dummycollege.org}}
\authorrunning{T. Maier and P. Sanders} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

@@

\begin{abstract}
Research on hash table data-structures has concentrated on optimizing
the speed and space efficiency for a long time.  Modern open
addressing techniques can fill a given table to more than 95\% before
the performance deteriorates significantly.  But once the threshold is
met, copying and migrating the whole table still seems to be the most
efficient answer.
Growing by a constant factor $\gamma$ often leads
to bad space efficiency because the newly grown table is only filled
less than a factor of $\gamma^{-1}$.

The space efficiency is even worse during the migration, where
oftentimes both the new and the old table coexist at the same time
causing a massive space overhead.  To solve both of these issues, we
propose the a new hash table architecture that supports dynamic
growing while keeping a strict size constraint $\alpha \cdot n$
throughout its lifetime.  Even under these conditions, we can
guarantee constant lookup times, and amortized insertion times,
independent of $n$.

We achieve this goal by splitting the hash table into subtables that
grow independent from each other.  The cuckoo load balancing
algorithm can move elements from one table to the other.  This makes
sure that growing one subtable increases insertion opportunities even
for elements that are not hashed into this one table.
 \end{abstract}

@@

\section{Introduction}
Hash tables are some of the most frequently used data-structures. They
belong into every programmers basic toolbox.  As such having
interchangeable implementations which perform well under different
conditions is important and allows programmers
easily adapt known solutions to new circumstances.

One aspect which a lot of research focuses on is space efficiency.
Modern space efficient hash tables work well up to loads of 95\% and
more. But for this to matter, programmers have to know similarly tight
bounds to the absolute number of elements inserted into the table.
This is typically not the case, one of the most ubiquitous use cases
for hash tables is to unify data elements by their key.  To
guarantee good performance, we have to overestimate the necessary
capacity whenever the exact number of unique keys is not known a
priori.  Dynamic data-structures are necessary to alleviate this
problem.  In circumstances where space efficiency is important, space
efficiently growing hash tables are necessary.

@@

To avoid both of these pitfalls we propose a variant of bucket cuckoo
hashing (TODO REFERENCE).  Bucket cuckoo hashing is a technique where
each element can be stored in one of several associated constant sized
buckets.  When all of them are full we move elements, to create new
space.  To solve the problem of efficient migration, we split one
bucket cuckoo table into multiple subtables each of which can grow
independently from all others.  Because the buckets associated with
one element are spread over the different subtables, growing one
subtable alleviates pressure from all others by allowing moves from a
dense table to the grown table.

@@

\section{Motivation}
\label{sec:motivation}
Space efficiency is a major factor for the development of new hashing
techniques.  The goal has long been to make hash tables work better in
high load scenarios.  Modern hashing techniques like Hopscotch, Robin
Hood, and Cuckoo Hashing can easily fill hash tables to more than 90\%
and up to 98\% while still achieving constant average running times on
most of their operations.  But since these techniques depend on
statically sized tables this only makes sense, if the number of unique
elements can be estimated with similar tight bounds.

To visualize this assume the following scenario.  During the operation
of a word count benchmark, we know an upper bound $c$ to the number of
unique words. Therefore, we construct a table large enough to hold
these elements ($0.9^{-1}\cdot c$ cells).  It is entirely possible
that the specific word count instance only contains $0.7\cdot c$
unique words (or less).  The resulting table is therefore only filled
63\%.  It is clear that better static sized tables cannot solve this
issue.

Depending on the underlying problem it is impossible to tightly
estimate the final number of elements.  Dynamic growing can be a
solution even in these instances.  It is completely independent from
the problem's domain and thus also works in cases where estimating the
number of elements can be hard.  To achieve space efficiency in
scenarios where the final size is not known the hash table has to grow
closely with the actual number of elements.  This is not efficient
with any of the current techniques used for hashing and migration.

\section{Related Work}
The usage of hash tables, and other hashing based algorithms has a
long history in computer science.  The classical methods results can
be found in all major algorithm textbooks (TODO ...).

Over the last one and a half decades, the field has regained
attention, both from theoretical and the practical point of view.  The
initial point of innovation was the power of two
choices~\cite{ThePowerOfTwoChoicesInRandomizedLoadBalancing}.
This concept led to the development of cuckoo hashing by Pagh and
Rodler~\cite{CuckooHashing}.  This technique reopened research into
more space efficient hash tables.  It proofed theoretically
interesting, because probabilistic bounds for the maximum fill degree
are highly non-trivial TODO SOME BOUNDS PAPERS.

Cuckoo hashing spawned many variants, that further improve its
performance.  The most notable one is probably bucket cuckoo hashing
($d$-ary cuckoo hashing), a variant that groups cells together in
buckets thus improving cache usage, and increasing displacement
opportunities TODO BUNCH OF PAPERS.  This further improves the maximum
fill degree. Nowadays, bucket cuckoo hashing is by far the most common
variant. There are even further variations on bucket cuckoo hashing,
like the generalization to continuous buckets.

@@

Some non-cuckoo space efficient hash tables continue to use linear
probing variants.  Robin Hood hashing is a technique that was
originally introduced in 1985 by Celis et
al.~\cite{RobinHoodHashing}. It has regained some popularity in
recent years -- mainly for its interesting theoretical
properties and the possibility to reduce the inherent variance of
linear probing.  Hopscotch~\cite{HopscotchHashing} hashing is a
technique, that originated in the area of concurrent hash tables, but
it can also lead to some densely filled hash tables.  For this reason
it is often used as a technique to achieve space efficient hash
tables.  This is only possible in a small number of circumstances, as
the stored bitmaps used in hopscotch hashing produce a significant
memory overhead.

@@

\section{Preliminaries}
A hash table is a data-structure, which stores key-value-pairs
($\langle key, data \rangle$) and offers the following functions:
\verb~insert~ -- stores a given key-value pair; \verb~find~ -- given a
key returns if said key was stored, and if so it returns the stored
value; and \verb~erase~ -- deletes a previously inserted element (if
present).

@@

Throughout this paper, we use $n$ to denote the number of elements and
$m$ the number of cells ($m > n$) in any given hash table.  We define
the load factor as $load = n/m$, and we use $\epsilon = 1-load$ for
the fraction of available cells.  Tables can usually only work
efficiently up to a certain load factor.  Above that, operations get
slower and/or have a possibility to fail.  When implementing a hash
table one has to decide between storing elements directly in the hash
table -- \emph{Closed Hashing} -- or storing pointers to elements --
\emph{Open Hashing}. This has an immediate impact on the necessary
memory.

\begin{align*}
  mem_{closed}(n,m) &\geq^* m\cdot size(element)\\
  mem_{open}  (n,m) &\geq^* m\cdot size(pointer) + n\cdot size(element)\\
\end{align*}

For large elements ($\gg size(pointer)$), one can simply use open
hashing, to reduce the relevant memory factor.  Therefore, we restrict
ourselves to the common case of elements whose size is comparable to
that of a pointer.  For our experiments we use 128bit elements (64bit
keys and 64bit values).  In these instances closed addressing
introduces a significant memory overhead (at least $\times1.5$),
therefore, we only consider closed hash tables for the purpose of this
paper. For closed hash tables, the memory efficiency is directly
dependent on the element density.  To reach high fill degrees with
closed hashing tables, we have to employ \emph{open addressing}
techniques which means, that elements are not stored in predetermined
cells, but can be stored in one of several options (e.g. linear
probing, or cuckoo hashing).

\subsection{$\alpha$-Space Efficient Hash Tables}
\label{sec:pre_staticspace}
We call a hashing technique $\alpha$-space efficient when it can work
effectively using at most $\alpha \cdot size_{\min} + O(1)$ memory. In
this case we define working efficiently as having average insertion
times close to $O(\frac{1}{\varepsilon})$. This is a natural bound
since it is the expected number of fully random probes needed to hit
an empty cell.

\[\alpha \cdot size_min = \alpha\cdot n\cdot size(element)\]

In many closed hashing techniques (e.g.~linear probing, cuckoo
hashing) cells are the same size as elements. Therefore, being
$\alpha$ space efficient is the same as operating with a load factor
of $\alpha^{-1} = n/m$ (because $\alpha\cdot n \cdot size(element) = m\cdot
size(element)$).  Because of this, we will mostly talk about the load
factor of a table instead of its memory usage.

@@

\paragraph*{$\alpha$-Space Efficiency for Dynamic Tables}
The definition of a space efficient hashing technique given aboce is
specifically targeted for statically sized hash tables.  As detailed
in Section~\ref{sec:motivation} our goal is to construct a hash table
which can efficiently (and tightly) grow with the number of inserted
elements. Therefore, we want to define the notion of an $\alpha$-space
efficient dynamic table.

We call a hash table implementation dynamically $\alpha$-space
efficient if an instantiated table can grow arbitrarily large over its
original capacity, while remaining smaller than $\alpha\cdot
mem_{necessary}$ at all times.
\[mem_{curr} \leq \alpha\cdot max(n_{curr})\cdot size(element)\]

@@

Similar to static $\alpha$-space efficiency, we will mostly talk about
the \emph{minimum load factor}, instead of $\alpha$.  It should be
noted, that this definition does not enforce size reduction due to the
deletion of elements, but it enforces that the cells used by deleted
elements must be reused.

\subsection{Cuckoo Hashing}
Cuckoo hashing is a technique to resolve hash conflicts in a closed
hashing hash table. Its main draw is that it guarantees constant
lookup times even in densely filled tables.  The distinguishing
technique of cuckoo hashing is, that $H$ hash functions ($h_1, ... ,
h_H$) are used to find $H$ independent possible positions. Each
element is stored in one of its positions.  Even if all possible
positions are occupied one can often move elements to create space for
the current element. We call this process displacing elements.

@@

\emph{Find} and \emph{remove} operations have a guaranteed constant
running time. Independent from the tables density, there are $H$
possible buckets -- $H\cdot B$ cells -- that have to be searched to
find an element (elements can be removed easily).

During an \emph{insert} the element is hashed to $H$ buckets.  If at least
one of those buckets has space left, then we store the element in the
bucket, that has the most space.  But when all buckets are full, then we
have to move elements around the table, such that a new space becomes
available.

@@

\begin{itemize}
\item copy some performance bounds from a cuckoo hashing paper
\item bla how dense can we get
\end{itemize}

When randomly probing the table we expect to find a free cell after
$\frac{1}{\varepsilon}$ random probes.

\section{DySECT Space Efficient Growing, by Splitting the Table}
Our goal is to build an efficiently growing data structure, that
remains space efficient at all times.  A commonly used growing
technique is to double the size by migrating all elements
into a larger new table.  Of course this is
not memory efficient.  A natural thought which occurs is to double
only part of the data structure.  This idea together with the
balancing ability of multiple hash functions is what will achieve the
described functionality.

\subsection{Description}
We use multiple ($T$) subtables (shown in TODO FIGREFRENCE) to allow
differentiated growing between table parts.  The subtables consist of
buckets that store elements.

Comparable to classic cuckoo hashing, each element has $H$ associated
buckets ($H$ number of hash functions), these can be in the same, or
in different subtables.  Each inserted element lies in one of its
associated buckets.  To find a bucket associated with an element $e$,
we compute $e$'s hash value using the appropriate hash function
$h_i(e)$, from a hashed key, we have to compute the subtable and the bucket
within that subtable.

To make this efficient we use powers of two for the number of
subtables ($T = 2^t$), as well as the individual subtable sizes ($s =
2^x$).  Since the number of subtables remains constant we can use the
first $t$ bits from the hashed key, to find the appropriate subtable.
From the remaining bits we compute the offset into this subtable using
a bitmask ($h_i(e)~\texttt{AND}~s-1 \leftrightarrow h_i(e) \mod s$).

@@

\subsection{Growing}
We allocate grow one subtable,
when the table contains enough elements, that the memory constraint
can be kept during the migration.  We migrate subtables in order
from the first to the last, therefore, no subtable can be more than
twice as large as any other.

Assume that all subtables have $s=m/T$ cells. When $\alpha\cdot n >
m+2s$ we can grow the first subtable while keeping the size
constraint.  Doubling the size of a subtable increases the global
number of cells from $m_{old} = T\cdot s$ to $m_{new} = m_{old}+s =
(T+1)\cdot s$ (factor: $\frac{T+1}{T}$).  Note that all subsequent
grows migrate one of the smaller tables, until all tables have the
same size again.  Therefore, each grow until then increase the
capacity by the same absolute amount (smaller factor).

@@

\subsection{Implementation Details}
For our experiments in Section~\ref{sec:exp} we use three hash
functions ($H=3$) and a bucket size of ($B=8$). These values have
consistently outperformed other possibilities.  We chose to use
$T=256$ subtables for all our tests.  The number of subtables only
plays a minor role for the running time.  We use breadth first search
as displacement strategy. Breadth first search performs better in
bucket cuckoo hashing, since it better uses the read cache lines from
one bucket.

@@

The added advantage is that we can grow subtables in-place. To map
virtual to physical memory we have to initialize part of the table. To
increase the size of a subtable, it is enough to initialize a new
section of the table.  Once this is done, we have to redistribute all
elements between the two parts of the table (moving only half the
elements).  This allows us to grow a subtable without the space
overhead of reallocation.

This way, we can also implement dynamic space efficient variants of
several other hashing schemes (see Section~\ref{sec:exp_competitors}).
These implementations, can easily compete with similar implementations
that use reallocation.

\subsection{Difficulties for the Analysis of our table structure}
There are three factors, that impact the performance of our dynamic
table compared to other cuckoo table variants and to other hashing solutions in
general \textbf{inhomogeneous table resolution}, \textbf{element imbalance}, and
\textbf{population density}. All of these factors influence the maximum load
density and the running times in different ways.

\subsubsection*{Imbalance through Inhomogeneous Table Resolution}
\label{sec:inhom_res}
By growing subtables individually we introduce a size imbalance
between subtables.  Large tables contain more buckets, but the number
of elements hashed to the table itself is not dependent on its size,
therefore, it is difficult to spread elements evenly among buckets.
Uneven bucket fill ratios can lead to longer insertion times.

@@

For two hash functions ($H=2$) and only one grown table ($j=1$) this
means that $\approx 2n/(T+1)$ should be stored in the first table, to
achieve a balanced bucket distribution.  Therefore, nearly all
elements associated to a bucket in the first table ($\approx 2n/T$)
have to be stored there.

\subsubsection*{Imbalance through Size Changes}
In addition to the problem of inhomogeneous tables, there is an
inherent balancing problem introduced by resizing subtables. It is
clear, that a newly grown table is not filled as densely as other
tables.  Since we double the table size, grown tables can only be
filled to about 50\%.

@@

\subsubsection*{Population Density}
Measuring and comparing the performance of space efficient growing
tables to their statically sized counterparts is difficult.  Inserting
elements into a densely filled table takes longer, than into an empty
one.  Therefore, a table that is always densely filled will naturally
be slower than one that only fills up towards the end of the
execution -- even when ignoring eventual growing costs.

@@

\subsubsection*{Comparison Implementations}
\label{sec:exp_competitor}
To put the performance of our solution into perspective, we
implemented and tested several other options for space efficient
hashing.  We use our own implementations, since most hash tables found
online do not support strict space-efficiency.  In Appendix TODO we
show that our implementations are at least competitive with
implementations found .

\paragraph*{Linear Probing} is the most common hashing technique outside of
hashing with chaining.  It is commonly accepted to be inefficient for
higher load factors because search distances can become large.  In
practice, this fact is counteracted by its cache-efficiency making
operations fast in spite of long probing distances.  Linear probing is
especially slow in \verb~find~ and \verb~erase~ operations, that often have
guaranteed constant running time in other implementation.

To grow our linear probing table, we use full table migrations.  This
can be made space efficient, by increasing the capacity in place.
This is possible due to over allocation (similar to
Section~\ref{sec:dys_inplace}) and a simple in place migration
algorithm (APPENDIX OR SECTION?).  For a table that was
initialized with $m$ cells with a min load factor $1-\varepsilon$ we
trigger growing once the table is loaded more than
$1-\frac{\varepsilon}{2}$. We then increase the capacity $m$ to
$(1-\varepsilon)^{-1}\cdot n$ to remain strictly size-efficient.  Our implementation
of the migration is highly cache efficient, but all elements have to
be moved.  This makes repeated migrations inefficient especially for
small $\alpha$.

\paragraph*{Hopscotch Hashing} is a technique similar to linear probing where
elements are moved when they are too far from their original hash
position.  Additionally, hopscotch hashing introduces an acceleration
data structure which uses bitmaps to store neighborhood data. This
data improves find and erase performance.  Many people would expect
the acceleration data to be the main draw to hopscotch hashing, but it
is also its greatest demise in the space-efficient setting.  The
bitmaps introduce a significant memory overhead of neighborhood-size
bits per cell.  Our experiments show that for any reasonable fill
factor neighborhoods have to be 64bits and larger, this introduces a
size factor of $\times1.5$ (128bit elements).  We still include
measurements with hopscotch hashing -- without considering space for
the neighborhood data -- to show its overall performance tendencies.
Growing hopscotch tables is done similar to the technique described
under linear probing.

@@

\paragraph*{Robin Hood Hashing} is another technique based on linear probing. The
goal of this technique is to improve find performance by minimizing
the maximum/average displacement distance.  The drawback of this
technique is that insertions have to compute the hash of each element
that is skipped by the insertion.  Migrating a Robin Hood hash table
is very fast, since the additional hash computations are not
necessary, and the element order remains the same (sorted by hash value).

>>>>

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{DySECT -- truly space efficient dynamic hashing} % -- a novel approach to growing!}
\titlerunning{A Sample LIPIcs Article} %optional, in case that the title is too long; the running title should fit into the top page column

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
\author[1]{Tobias Maier}
\author[1]{Peter Sanders}
\affil[1]{Karlsruhe Institute of Technology, Karlsruhe, Deutschland\\
  \texttt{\{t.maier, sanders\}@kit.edu}}
%% \affil[2]{Department of Informatics, Dummy College, Address/City, Country\\
%%   \texttt{access@dummycollege.org}}
\authorrunning{T. Maier and P. Sanders} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

@@

\begin{abstract}
Research has concentrated on optimizing the speed and space efficiency
of hash tables for a long time.  Modern open addressing techniques can
fill a given table to more than 95\% before the performance
deteriorates significantly.  But once the threshold is met, copying
and migrating the whole table still seems to be the most efficient
answer.  Growing by a constant factor $\gamma$ often leads to bad
space efficiency because the newly grown table is only filled less
than a factor of $\gamma^{-1}$.

The space efficiency is even worse during the migration, where
oftentimes both the new and the old table coexist at the same time
causing a massive space overhead.  To solve both of these issues, we
propose the a new hash table architecture that supports dynamic
growing while keeping a strict size constraint $\alpha \cdot n$
throughout its lifetime.  Even under these conditions, we achieve
guaranteed constant lookup times, and amortized insertion times,
independent of $n$.

We achieve this goal by splitting the hash table into subtables that
grow independent from each other.  The cuckoo load balancing mechanism
can move elements from one table to the other.  This ensures that
growing one subtable increases insertion opportunities even for
elements that are not hashed into this one table.
 \end{abstract}

@@

\section{Introduction}
Hash tables are some of the most frequently used data-structures. They
belong into every programmers basic toolbox.  As such having
interchangeable implementations which perform well under different
conditions is important and allows programmers to easily adapt known
solutions to new circumstances.

One aspect which a lot of research focuses on is space efficiency.
Modern space efficient hash tables work well up to loads of 95\% and
more. But for this to matter, programmers have to know similarly tight
bounds to the absolute number of elements inserted into the table.
This is typically not the case, one of the most ubiquitous use cases
for hash tables is to unify data elements by their key.  To guarantee
good performance, we have to overestimate the initial capacity
whenever the exact number of unique keys is not known a priori.
Dynamic data-structures are necessary to alleviate this problem.  In
circumstances where space efficiency is important, we need space
efficiently growing hash tables.

@@

To avoid both of these pitfalls we propose a variant of bucket cuckoo
hashing (TODO REFERENCE).  Bucket cuckoo hashing is a technique where
each element can be stored in one of several associated constant sized
buckets.  When all of them are full we move elements, to create new
space.  To solve the problem of efficient migration, we split the
table into multiple subtables each of which can grow independently
from all others.  Because the buckets associated with one element are
spread over the different subtables, growing one subtable alleviates
pressure from all others by allowing moves from a dense table to the
grown table.

@@

\section{Motivation}
\label{sec:motivation}
Space efficiency is a major factor for the development of new hashing
techniques.  The goal has long been to make hash tables work better in
high load scenarios.  Modern hashing techniques like Hopscotch, Robin
Hood, and Cuckoo Hashing can easily fill hash tables to more than 90\%
and up to 98\% while still achieving constant average running times on
most of their operations.  But since these techniques depend on
statically sized tables this only makes sense, if the number of unique
elements can be estimated with similarly tight bounds.

To visualize this assume the following scenario.  During the operation
of a word count benchmark, we know an upper bound $c$ to the number of
unique words. Therefore, we construct a table large enough to hold
these elements ($0.9^{-1}\cdot c$ cells).  It is entirely possible
that the specific word count instance only contains $0.7\cdot c$
unique words (or less).  The resulting table is therefore only filled
63\%.  It is clear that better static sized tables cannot achieve fill
ratios greater than 70\%.

Depending on the underlying problem it is impossible to tightly
estimate the final number of elements. Since dynamic growing is independent from the problem's domain, it can be a
solution even in these instances.  To achieve space efficiency in
scenarios where the final size is not known the hash table has to grow
closely with the actual number of elements.  This is not efficient
with any of the current techniques used for hashing and migration.

\section{Related Work}
The usage of hash tables, and other hashing based algorithms has a
long history in computer science.  The classical methods and results
can be found in all major algorithm textbooks (TODO ...).

Over the last one and a half decades, the field has regained
attention, both from theoretical and the practical point of view.  The
initial innovation that sparked this attention was the power of two
choices~\cite{ThePowerOfTwoChoicesInRandomizedLoadBalancing}.  This
concept led to the development of cuckoo hashing~\cite{CuckooHashing}
which reopened research into more space efficient hash tables.  They
proofed theoretically interesting, because probabilistic bounds for
the maximum fill degree are often highly non-trivial TODO SOME BOUNDS
PAPERS.

Cuckoo hashing also spawned many variants, that further improve its
performance.  The most notable one is probably bucket cuckoo hashing
($d$-ary cuckoo hashing), a variant that groups cells together in
buckets thus improving cache usage, and increasing displacement
opportunities TODO BUNCH OF PAPERS.  This further improves the maximum
fill degree. Nowadays, bucket cuckoo hashing is by far the most common
variant. There are even further variations on bucket cuckoo hashing,
like the generalization to continuous buckets TODO PAPER.

@@

Some non-cuckoo space efficient hash tables continue to use linear
probing variants.  Robin Hood hashing is a technique that was
originally introduced in 1985~\cite{RobinHoodHashing}. It has regained some popularity in
recent years -- mainly for its interesting theoretical
properties and the possibility to reduce the inherent variance of
linear probing.  Hopscotch~\cite{HopscotchHashing} hashing is a
technique, that originated in the research of concurrent hash tables, but
it can also lead to some densely filled hash tables.  For this reason
it is often used as a technique to achieve space efficient hash
tables.  This is only possible in a small number of circumstances, as
the stored bitmaps used in hopscotch hashing create a significant
memory overhead.

@@

\section{Preliminaries}
A hash table is a data-structure, which stores key-value-pairs
($\langle key, data \rangle$) and offers the following functions:
\verb~insert~ -- stores a given key-value pair; \verb~find~ -- given a
key returns if said key was stored, and $\bot$ otherwise; and
\verb~erase~ -- deletes a previously inserted element (if present).

@@

Throughout this paper, we use $n$ to denote the number of elements and
$m$ the number of cells ($m > n$) in any given hash table.  We define
the load factor as $load = n/m$, and we use $\varepsilon = 1-load$ for
the fraction of available cells.  Tables can usually only work
efficiently up to a certain load factor.  Above that, operations get
slower and/or have a possibility to fail.  When implementing a hash
table one has to decide between storing elements directly in the hash
table -- \emph{Closed Hashing} -- or storing pointers to elements --
\emph{Open Hashing}. This has an immediate impact on the necessary
memory.

\begin{align*}
  mem_{closed}(n,m) &\geq m\cdot size(element)\\
  mem_{open}  (n,m) &\geq m\cdot size(pointer) + n\cdot size(element)\\
\end{align*}

For large elements ($\gg size(pointer)$), one can use a non-space
efficient hash table with open hashing, to reduce the relevant memory
factor.  Therefore, we restrict ourselves to the common and more interesting case of
elements whose size is comparable to that of a pointer.  For our
experiments we use 128bit elements (64bit keys and 64bit values).  In
these instances open hashing introduces a significant memory
overhead (at least $\times1.5$), therefore, we only consider closed
hash tables for the purpose of this paper. For closed hash tables, the
memory efficiency is directly dependent on the element density.  To
reach high fill degrees with closed hashing tables, we have to employ
\emph{open addressing} techniques which means, that elements are not
stored in predetermined cells, but can be stored in one of several
options (e.g. linear probing, or cuckoo hashing).

\subsection{$\alpha$-Space Efficient Hash Tables}
\subparagraph*{Static}
\label{sec:pre_staticspace}
We call a hashing technique $\alpha$-space efficient when it can work
effectively using at most $\alpha \cdot n_{\textit{curr}} \cdot \textit{size}(\textit{element}) + O(1)$ memory. In
this case we define working efficiently as having average insertion
times close to $O(\frac{1}{\varepsilon})$. This is a natural bound
since it is the expected number of fully random probes needed to hit
an empty cell.

%\[\alpha \cdot size_min = \alpha\cdot n\cdot size(element)\]

In many closed hashing techniques (e.g.~linear probing, cuckoo
hashing) cells are the same size as elements. Therefore, being
$\alpha$ space efficient is the same as operating with a load factor
of $\alpha^{-1} = n/m$.
%(because $\alpha\cdot n \cdot size(element) = m\cdot size(element)$).
Because of this, we will mostly talk about the load factor of a table
instead of its memory usage.

@@

\paragraph*{Dynamic}
The definition of a space efficient hashing technique given aboce is
specifically targeted for statically sized hash tables.  We call a
hash table implementation dynamically $\alpha$-space efficient if an
instantiated table can grow arbitrarily large over its original
capacity, while remaining smaller than $\alpha\cdot n_{\max}\cdot
\textit{size}(\textit{element}) + O(1)$ at all times.
%\[mem_{curr} \leq \alpha\cdot max(n_{curr})\cdot size(element)\]

@@

Similar to static $\alpha$-space efficiency, we will mostly talk about
the \emph{minimum load factor}, instead of $\alpha$.  We should note,
that this definition of dynamic $\alpha$-space efficient hash tables
does not enforce size reduction due to the deletion of elements, but
it enforces that cells previously used by deleted elements must be reused.

\subsection{Cuckoo Hashing}
\label{sec:pre_cuckoo}
Cuckoo hashing is a technique to resolve hash conflicts in a closed
hashing hash table. Its main draw is that it guarantees constant
lookup times even in densely filled tables.  The distinguishing
technique of cuckoo hashing is, that $H$ hash functions ($h_1, ... ,
h_H$) are used to compute $H$ independent positions. Each element is
stored in one of its positions.  Even if all positions are occupied
one can often move elements to create space for the current
element. We call this process displacing elements.

@@

\emph{Find} and \emph{remove} operations have a guaranteed constant
running time. Independent from the tables density, there are $H$
buckets -- $H\cdot B$ cells -- that have to be searched to find an
element.

During an \emph{insert} the element is hashed to $H$ buckets.  If at least
one of those buckets has an empty cell, we store the element in the
bucket with the most space.  When all buckets are full, then we
have to move elements around the table, such that a new space becomes
available.

@@

\section{DySECT Space Efficient Growing, by Splitting the Table}
Our goal is to build an efficiently growing data structure, that
remains space efficient at all times.  A commonly used growing
technique is to double the size by migrating all elements
into a larger new table.  Of course this is
not memory efficient.  A natural thought is to double
only part of the data structure.  This idea together with the
balancing mechanism of cuckoo hashing will achieve the
described functionality.

\subsection{Description}
We use $T$ subtables (shown in TODO FIGREFRENCE) to allow
differentiated growing between table parts.  The subtables consist of
buckets that store elements.

Each element has $H$ associated
buckets similar to cuckoo hashing, these can be in the same, or
in different subtables.  To find a bucket associated with an element $e$,
we compute $e$'s hash value using the appropriate hash function
$h_i(e)$, from a hashed key, we have to compute the subtable and the bucket
within that subtable.

To make this efficient we use powers of two for the number of
subtables ($T = 2^t$), as well as the individual subtable sizes ($s =
2^x*B$).  Since the number of subtables is constant we can use the
first $t$ bits from the hashed key, to find the appropriate subtable.
From the remaining bits we compute the offset into this subtable using
a bitmask ($h_i(e)\,\texttt{\&}\,(s-1) = h_i(e) \mod s$).

@@

\subsection{Growing}
We grow one subtable,
when the table contains enough elements, that the memory constraint
can be kept during the migration.  We migrate subtables in order
from the first to the last, therefore, no subtable can be more than
twice as large as any other.

Assume that all subtables have $s=m/T$ cells. When $\alpha\cdot n >
m+2s$ we can grow the first subtable while keeping the size
constraint.  Doubling the size of a subtable increases the global
number of cells from $m_{old} = T\cdot s$ to $m_{new} = m_{old}+s =
(T+1)\cdot s$ (grow factor $\frac{T+1}{T}$).  Note that all subsequent
grows migrate one of the smaller tables, until all tables have the
same size again.  Therefore, each grow until then increase the
capacity by the same absolute amount (smaller grow factor).

@@

\subsection{Implementation Details}
For our experiments in Section~\ref{sec:exp} we use three hash
functions ($H=3$) and a bucket size of ($B=8$). These values have
consistently outperformed other options.  We chose to use $T=256$
subtables for all our tests.  The number of subtables only plays a
minor role for the running time.  To find displacement opportunities
we use breadth first search.  In our tests it performed better than
random walks, since it better uses the read cache lines from one
bucket.

@@

The added advantage is that we can grow subtables in-place.  To
increase the size of a subtable, it is enough to initialize a
consecutive section of the table (following the original subtable).
Once this is done, we have to redistribute the table's elements.  This
allows us to grow a subtable without the space overhead of
reallocation.

This way, we can also implement in place growing variants of other
hashing schemes (see Section~\ref{sec:exp_competitors}).  These
implementations have the same performance of similar implementations
using full table migration.  But given the correct growing factor,
they can be space efficient.

\subsection{Difficulties for the Analysis of our table structure}
There are two factors, that impact the performance of our dynamic
table compared to other cuckoo table variants and to other hashing
solutions in general \textbf{inhomogeneous table resolution} and
\textbf{element imbalance}.
%, and \textbf{population density}.
All of these factors influence the maximum load density and the
running times in different ways.

\subparagraph*{Imbalance through Inhomogeneous Table Resolution}
\label{sec:inhom_res}
By growing subtables individually we introduce a size imbalance
between subtables.  Large tables contain more buckets, but the number
of elements hashed to the table itself is not dependent on its size,
therefore, it is difficult to spread elements evenly among buckets.
Uneven bucket fill ratios can lead to longer insertion times.

@@

For two hash functions ($H=2$) and only one grown table ($j=1$) this
means that $\approx 2n/(T+1)$ should be stored in the first table, to
achieve a balanced bucket distribution.  Therefore, nearly all
elements associated to a bucket in the first table ($\approx 2n/T$)
have to be stored there.  This is the reason, why $H=3$ leads to
significantly better space efficiency than $H=2$.

\subparagraph*{Imbalance through Size Changes}
In addition to the problem of inhomogeneous tables, there is an
inherent balancing problem introduced by resizing subtables. It is
clear, that a newly grown table is not filled as densely as other
tables.  Since we double the table size, grown tables can only be
filled to about 50\%.

@@

% \subsubsection*{Population Density}
% Measuring and comparing the performance of space efficient growing
% tables to their statically sized counterparts is difficult.  Inserting
% elements into a densely filled table takes longer, than into an empty
% one.  Therefore, a table that is always densely filled will naturally
% be slower than one that only fills up towards the end of the
% execution -- even when ignoring eventual growing costs.

@@

\subsubsection*{Comparison Implementations}
\label{sec:exp_competitors}
To put the performance of our solution into perspective, we
implement and test several other options for space efficient
hashing.  We use our own implementations, since no hash table found
online supports strict space-efficiency.

\subparagraph*{Linear Probing} is the most common hashing technique outside of
hashing with chaining.  It is commonly accepted to be inefficient for
higher load factors because search distances can become large.  In
practice, this fact is counteracted by its cache-efficiency making
operations fast in spite of long probing distances.  Linear probing is
especially slow in \verb~find~ and \verb~erase~ operations, that often have
guaranteed constant running time in other implementation.

To grow our linear probing table, we use full table migrations.  This
can be made space efficient, by increasing the capacity in place.
This is possible due to over allocation (similar to
Section~\ref{sec:dys_inplace}) and a simple in place migration
algorithm (APPENDIX OR SECTION?).  For a table that was
initialized with $m$ cells with a min load factor $1-\varepsilon$ we
trigger growing once the table is loaded more than
$1-\frac{\varepsilon}{2}$. We then increase the capacity $m$ to
$(1-\varepsilon)^{-1}\cdot n$ to remain strictly size-efficient.  Our implementation
of the migration is highly cache efficient, but all elements have to
be moved.  This makes repeated migrations inefficient especially for
small $\varepsilon$.

\subparagraph*{Hopscotch Hashing} is a technique similar to linear probing where
elements are moved when they are too far from their original hash
position.  Additionally, hopscotch hashing introduces an acceleration
data structure which uses bitmaps to store neighborhood data. This
data improves find and erase performance.  Many people would expect
the acceleration data to be the main draw to hopscotch hashing, but it
is also its greatest demise in the space-efficient setting.  The
bitmaps introduce a significant memory overhead of neighborhood-size
bits per cell.  Our experiments show that for any reasonable fill
factor neighborhoods have to be 64bits and larger, this introduces a
size factor of $\times1.5$ (128bit elements).  We still include
measurements with hopscotch hashing -- without considering space for
the neighborhood data -- to show its overall performance tendencies.
Growing hopscotch tables is done similar to the technique described
under linear probing.

@@

\subparagraph*{Robin Hood Hashing} is another technique based on linear probing. The
goal of this technique is to improve find performance by minimizing
the maximum displacement distance.  The drawback of this
technique is that insertions have to compute the hash of each element
that is skipped by the insertion.  Migrating a Robin Hood hash table
is very fast, since the additional hash computations are not
necessary, and the element order remains the same (sorted by hash value).

\subparagraph*{Classic Bucket Cuckoo Hashing} as described in
Section~\ref{sec:pre_cuckoo} is the basic hashing technique, that
inspired our hashing technique.  Like our variant it offers constant
lookup and deletion times.  But it suffers from an inefficient growing
technique similar to the one used by the linear probing table above.
Similar to DySECT, $H=3$ and $B=8$ have turned out to be an optimum
for performance.
